---
title: "w04_Task3"
---

## Load in Packages
```{r, message="FALSE", warning="FALSE"}

library(haven)     # read_dta() for Stata, read_sav() for SPSS
library(foreign)   # read.dbf() for the DBF file inside the zip
library(tidyverse) # filter, select, rename, etc.

```

## Load in Data
```{r, message="FALSE", warning="FALSE", results="hide"}


german_conscr <- read_dta(
  "https://byuistats.github.io/M335/data/heights/germanconscr.dta"
)


german_prison <- read_dta(
  "https://byuistats.github.io/M335/data/heights/germanprison.dta"
)



zip_path <- tempfile(fileext = ".zip")
download.file(
  url      = "https://byuistats.github.io/M335/data/heights/Heights_south-east.zip",
  destfile = "german_soldier.zip",
  mode     = "wb"          # binary mode — essential for zip files
)

unzip_dir <- tempdir()
unzip("german_soldier.zip", exdir = unzip_dir)
list.files(unzip_dir, pattern = "\\.dbf$", ignore.case = TRUE)

dbf_path <- file.path(unzip_dir, "B6090.DBF")
se_german <- read.dbf(dbf_path, as.is = TRUE)


bls_heights <- read_csv(
  "https://raw.githubusercontent.com/hadley/r4ds/main/data/heights.csv"
)


wisconsin <- read_sav(
  "https://www.ssc.wisc.edu/nsfh/wave3/NSFH3%20Apr%202005%20release/main05022005.sav"
)


```

## Wrangle Data
```{r}


# 1 - Bavarian conscripts
german_conscr_clean <- german_conscr |>
  mutate(
    birth_year = bdec,
    height.cm  = height,
    height.in  = height / 2.54,
    study      = "Bavarian Conscripts"
  ) |>
  select(birth_year, height.in, height.cm, study)

# 2 - Bavarian prison
german_prison_clean <- german_prison |>
  mutate(
    birth_year = bdec,
    height.cm  = height,
    height.in  = height / 2.54,
    study      = "Bavarian Prison"
  ) |>
  select(birth_year, height.in, height.cm, study)

# 3 - SE German soldiers
se_german_clean <- se_german |>
  mutate(
    birth_year = GEBJ,
    height.cm  = CMETER,
    height.in  = CMETER / 2.54,
    study      = "SE German Soldiers"
  ) |>
  select(birth_year, height.in, height.cm, study)

# 4 - BLS heights (inches -> cm, males only, assign birth year)
bls_clean <- bls_heights |>
  filter(sex == "male") |>
  mutate(
    birth_year = 1950,
    height.in  = height,
    height.cm  = height * 2.54,
    study      = "BLS Heights"
  ) |>
  select(birth_year, height.in, height.cm, study)

# 5 - Wisconsin
# DOBY is 2-digit year, RT216F = feet, RT216I = remaining inches
# Filter out impossible inch values (>11) and impossible heights overall
wisconsin_clean <- wisconsin |>
  mutate(
    birth_year = 1900 + DOBY,
    height.in  = (RT216F * 12) + RT216I,
    height.cm  = height.in * 2.54
  ) |>
  filter(
    RT216I <= 11,          # remove impossible inch remainders
    height.in >= 48,       # remove implausibly short (< 4 ft)
    height.in <= 96        # remove implausibly tall (> 8 ft)
  ) |>
  mutate(study = "Wisconsin NSFH") |>
  select(birth_year, height.in, height.cm, study)


```

## Combine Data into One Table
```{r}

heights_combined <- bind_rows(
  german_conscr_clean,
  german_prison_clean,
  se_german_clean,
  bls_clean,
  wisconsin_clean
)

```

## Data Wrangling Process

Each of the five datasets required new columns for birth_year, height.in, height.cm, and study to be created through mutation and renaming. The three German datasets stored height in centimeters, so height.in was derived by dividing by 2.54. The BLS dataset was already in inches, so height.cm was derived by multiplying by 2.54, and birth year was assigned as 1950 since none was provided. The Wisconsin dataset was the most complex, with height split across a feet column and a remaining-inches column that had to be combined, and birth year stored as a two-digit value requiring the addition of 1900. Because the German datasets represent exclusively male populations by their nature as military and prison records, the BLS dataset was filtered to male respondents only to keep comparisons consistent across studies.

```{r, message="FALSE", warning="FALSE", results="hide"}

study_labels <- heights_combined |>
  group_by(study) |> 
  summarise(
    min_yr = min(birth_year, na.rm = TRUE),
    max_yr = max(birth_year, na.rm = TRUE)
  ) |>
  mutate(label = ifelse(min_yr == max_yr,
                        paste0(study, "\n(", min_yr, ")"),
                        paste0(study, "\n(", min_yr, "–", max_yr, ")")))

label_vec <- setNames(study_labels$label, study_labels$study)


graph <- heights_combined |>
  mutate(study = fct_reorder(study, birth_year, .fun = median, na.rm = TRUE)) |>
  ggplot(aes(x = study, y = height.in, fill = study)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.08, fill = "white", outlier.shape = NA) +
  scale_x_discrete(labels = label_vec) +
  labs(
    title = "Height distribution across centuries", 
    x = NULL,
    y = "Height (inches)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "none")

```

## Conclusion

The previous task used aggregate world height data and told a fairly clean story -- humans have grown roughly 2.2 inches taller on average over the last 150 years, a modest but consistent upward trend visible across many countries and regions. The individual-level data in this task loosely supports that same story. Following the median lines in the violin plot left to right, there is a subtle upward drift from the German military datasets clustered around 67-68 inches toward the American datasets nudging 69-70 inches. However, these datasets are not a controlled comparison. Three studies draw from German and Bavarian military and prison populations, which had minimum height requirements that cut off the lower tail of the true distribution, and then the comparison jumps to American survey respondents entirely. Any visible difference could reflect geography, diet, or genetics just as easily as the passage of time.

The graph also reveals some honest limitations worth noting. The SE German Soldiers violin has a jagged, spiky shape suggesting thin or unevenly distributed data rather than a clean population sample. The Wisconsin violin is unusually wide and tall at the top, almost certainly a consequence of its mixed gender population pulling the distribution in two directions at once. BLS, assigned a single birth year of 1950, reflects cross-sectional variation rather than any temporal trend at all. Taking both tasks together, the aggregate world data makes a more convincing case for humans getting taller over time than this individual-level data does on its own. The trend appears real, but it is modest enough that selection bias and population differences across these five studies can easily obscure it.

### AI Use

AI assisted in this process (Claude Sonnet 4.6, Anthropic, May 2026), helping identify column names, diagnose data issues, and suggest wrangling decisions; all outputs were reviewed and validated by me.